---
title: "643-FP-ContentBased"
author: "James Topor"
date: "July 7, 2017"
output: html_document
---


```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(recommenderlab)
library(readr)
library(matrixStats)
library(knitr)
library(tidyr)
library(dplyr)
library(proxy)

# turn off scientific notation
options(scipen=999)

set.seed(1)
```


Load list of artists

```{r, message = FALSE, warning = FALSE}
# load list of last.fm artists: drop columns containing URL's since they aren't needed
lfm_art <- read_delim("https://raw.githubusercontent.com/jtopor/CUNY-MSDA-643/master/FP/artists.dat", delim = "\t") %>% select(id, name)

# cleanup foreign characters in artist names: most will be converted to '?'
lfm_art$name <- iconv(lfm_art$name, from = "UTF-8", to = "ASCII//TRANSLIT")
```


Load user-artists, find top 1000 artists (as determined by number of listeners), and remove any artists that don't have an entry in the artists.dat file

```{r}
# load last.fm user_artists file
lastfm <- read.table("https://raw.githubusercontent.com/jtopor/CUNY-MSDA-643/master/FP/user_artists.dat", header = TRUE, sep = "", stringsAsFactors = FALSE)

# calc number of users listening to each artist
a_users <- arrange(summarise(group_by(lastfm, artistID), 
                     TotalUsers = length(unique(userID)) ), desc(TotalUsers) )

# truncate at top 1000
top_1000 <- a_users[1:1000,]

# find names of artists with most last.fm fans
most_fans <- subset(top_1000, artistID %in% lfm_art$id)

# re-arrange sort order to enable proper link to artist name
most_fans <- arrange(most_fans, artistID, TotalUsers)

# get names of artists
mf_names <- subset(lfm_art, id %in% most_fans$artistID)

most_fans$Name <- mf_names$name[mf_names$id %in% most_fans$artistID]

most_fans <- arrange(most_fans, desc(TotalUsers))

missing <- subset(top_1000, !(artistID %in% most_fans$artistID))


# remove all items not in top 1000 artist list
last_sm <- subset(lastfm, artistID %in% top_1000$artistID)

# remove all artist ID's missing from artists.dat file
last_sm <- subset(last_sm, !(artistID %in% missing$artistID))

# form new master list of valid artist ID's excluding the 182 missing ones
top_818 <- subset(top_1000, !(artistID %in% missing$artistID))

rm(top_1000)
```


Load list of genres / tags:

```{r}
# load last.fm tags.dat file
lfm_tags <- read_delim("https://raw.githubusercontent.com/jtopor/CUNY-MSDA-643/master/FP/tags.dat", delim = "\t")

# count distinct users
length(unique(lfm_tags$tagID))

```

11,946 possible genres / tags



Load list of UserID / Tag pairs: 

```{r}
# load last.fm user-taggedartists.dat file
user_tags <- read_delim("https://raw.githubusercontent.com/jtopor/CUNY-MSDA-643/master/FP/user_taggedartists.dat", delim = "\t") %>% select(userID, artistID, tagID)

# count entries in file
nrow(user_tags)

# count distinct users
length(unique(user_tags$userID))

# count distinct artists
length(unique(user_tags$artistID))
```

186,479 entries

1892 users (which means all users are represented)

12,523 distinct artists have been tagged

_____

Summary stats on number of tags applied per user
```{r}
summary(summarise(group_by(user_tags, userID),
                     TotalTags = length(userID == userID) )$TotalTags )
```

So we have a median of 20 tags applied per user with a mean of 98.56.

_____

Summary stats for number of genres used as tags per user:
```{r}
summary(summarise(group_by(user_tags, userID),
                     TotalUTags = length(unique(tagID)) )$TotalUTags )
```

Median of 12 different genre tags used by each user to categorize an artist within a genre, with a mean of 18.93. 


Find most widely used tags:
```{r}
# calc number of users listening to each artist
tag_counts <- arrange(summarise(group_by(user_tags, tagID), 
                     TotalUsers = length(unique(userID)) ), desc(TotalUsers) )

summary(tag_counts$TotalUsers)
```

Most genre tags aren't used much: median = 1, mean = 3.674


```{r}
par(mfrow=c(1,2))
hist(tag_counts$TotalUsers, col = "yellow", main = "Dist. of # of Genre Taggings", breaks = 50, xlab = "Number of Listeners")

boxplot(tag_counts$TotalUsers, col = "yellow", main = "Dist. of # of Genre Taggings", ylab = "Number of Listeners")
```

Try truncating to top 200 genres

```{r}
# truncate at top 400
tag_200 <- tag_counts[1:200,]

tag_200 <- arrange(tag_200, tagID)

# get tag names
tag_200$Names <- subset(lfm_tags, tagID %in% tag_200$tagID)$tagValue

# sort by number of users
tag_200 <- arrange(tag_200, desc(TotalUsers))

kable(head(cbind(tag_200$Names, tag_200$TotalUsers), 20), col.names = c("Genre", "Num users Applying Tag"))
```

Create artist-genre matrix using user_tags + tag_200 data

First exclude tags that are not in top 200 from user_tags:

```{r}
u_toptags <- subset(user_tags, tagID %in% tag_200$tagID)

# count distinct artists
length(unique(u_toptags$artistID))
```


Now exclude artists not in top 818 list
```{r}
u_toptags <- subset(u_toptags, artistID %in% top_818$artistID)

# count distinct artists
length(unique(u_toptags$artistID))
```


Now ensure that artistID / tagID pairs are summarized by counting the number of times
any given tag has been applied to an artist:
```{r}
u_tt <- summarise(group_by(u_toptags, artistID, tagID ),
                       Count = length(tagID) )

# count distinct artists
length(unique(u_tt$artistID))

```

So we have a mismatch: 3 of the 818 artists we retained from the user-artists data have not been genre tagged by any user via the top 200 genre tags

Which 3?

```{r}
# get a list of artists that haven't been tagged with one of top 200 tags
not_tagged <- subset(top_818, !(artistID %in% u_toptags$artistID))
not_tagged # all have relatively low user counts so OK to discard

# check to see whether artists have been tagged at all
not_tagged$artistID %in% user_tags$artistID

# they have been tagged, but not with one of top 200 tags

```


Remove 3 non-tagged artists from top_818

```{r}
top_815 <- subset(top_818, artistID %in% u_toptags$artistID)
# rm(top_818)

# count distinct artists
length(unique(top_815$artistID))

```


Remove artists not tagged using top 200 tags from user-artists data

```{r}
# remove all artist ID's missing from artists.dat file
last_sm <- subset(last_sm, artistID %in% top_815$artistID)

# count distinct users
length(unique(last_sm$userID))

```

1870 users retained - same as with top 818 artists

_____

# Create User-Artist Matrix

```{r}
# convert to wide format
l_mat <- spread(last_sm, artistID, weight)

# save UserIDs and remove 1st column from matrix
user_ids <- as.vector(l_mat$userID)

# create a matrix using cols 2 -> ncol of l_mat
lr_mat <- as.matrix(l_mat[,2:ncol(l_mat)])
```

Calculate density / sparsity

```{r}
# calc number of ratings in matrix
nratings <- length(as.vector(lr_mat))
nratings

# calc density of matrix = 0.0337
sum(!is.na(as.vector(lr_mat)) ) / nratings

# calc sparsity of matrix
1 - sum(!is.na(as.vector(lr_mat)) ) / nratings
```

_____

# Create Artist - Genre Matrix

```{r}
# convert to wide format
tmp_mat <- spread(u_tt, tagID, Count)

# save artistIDs and remove 1st column from matrix
ag_artistID <- as.vector(tmp_mat$artistID)

# create a matrix using cols 2 -> ncol of l_mat
ag_mat <- as.matrix(tmp_mat[,2:ncol(tmp_mat)])

rm(tmp_mat)
```

Do a data integrity check to ensure content of artist-genre matrix is valid
relative to original user-tags data

```{r}
# sanity check:
# artistID 56 has been tagged with genre 18 39 times
nrow(subset(user_tags, artistID == 56 & tagID == 18))

# artistID 72 has been tagged with genre 18 68 times
nrow(subset(user_tags, artistID == 72 & tagID == 18))

# artistID 89 has been tagged with genre 24 99 times
nrow(subset(user_tags, artistID == 89 & tagID == 24))
```

Calculate density / sparsity

```{r}
# calc number of ratings in matrix
ntags <- length(as.vector(ag_mat))
ntags

# calc density of matrix = 0.091
sum(!is.na(as.vector(ag_mat)) ) / ntags

# calc sparsity of matrix
1 - sum(!is.na(as.vector(ag_mat)) ) / ntags
```

_____

# Summary

- The user-artists matrix is named lr_mat

- The artist-genre matrix is named ag_mat

- A matching set of 815 artists is used in each matrix, thus ensuring we can tie any artist within user-artists to the genres with which they've been tagged

- 1870 total users represented in user-artists

- user-artists contains 1,524,050 possible ratings, with a density of __3.37%__

- artists-genre contains 163,000 possible genre taggings, with a density of __9.1%__

- The user-item matrix can be binarized as follows:

```{r, eval = TRUE}
# create binarized copy of data
bin_lrmat <- lr_mat

bin_lrmat[,][is.na(bin_lrmat[,])] <- 0
bin_lrmat[,][bin_lrmat[,] > 0] <- 1
```


- The artist-genre matrix can be binarized as follows:

```{r, eval = TRUE}
# create binarized copy of data
bin_agmat <- ag_mat

bin_agmat[,][is.na(bin_agmat[,])] <- 0
bin_agmat[,][bin_agmat[,] > 0] <- 1
```

_____ 

# Build a User-Based Collaborative Filter

Create training + Test Set using binary user-artist data

```{r}
# raw_median <- median(as.vector(bin_lrmat), na.rm = TRUE)

# convert non-binary matrix to a recommenderlab realRatingMatrix
ua_bmat <- as(bin_lrmat,"binaryRatingMatrix")
```

Build the model + run once

```{r}
# split the binary data into the training and the test set:
e_bin <- evaluationScheme(ua_bmat, method="split", train=0.8, given = 1, goodRating = 1)

n_recommended <- 10

# build the item-based binary recommender using training subset
b1 <- Recommender(getData(e_bin, "train"), "UBCF", 
                          parameter = list(method = "Jaccard"))

# make predictions on test set
b_pred <- predict(b1, getData(e_bin, "known"), n = n_recommended, goodRating = 1)

# check the accuracy of the predictions
error_b <- calcPredictionAccuracy(b_pred, getData(e_bin, "unknown"), 
                                  given = n_recommended, goodRating = 1)

kable(error_b, caption = "Performance Metrics")
```

As we can see below, the system does appear to be generating a list of 10 recommended artist ID's per user:

```{r}
b_pred@items[1:4]
```


Make Top 10 recs for EVERY user. First step is get top 20
```{r}
n_recommended <- 20

# now make predictions for every user with the binary recommender
b_pred <- predict(b1, ua_bmat, n = n_recommended, goodRating = 1)

# check to ensure rec's created for all users
b_pred@items[20:23]
```

Create data frame to store top 10 recs for all users, ensuring that all have at 
least 10

```{r}
# create a data frame to house 10 recommendations for each artist
user_tenrecs <- data.frame(matrix(ncol = 11, nrow = length(user_ids)))
user_tenrecs[,1] <- user_ids


colnames(user_tenrecs) <- c("userID", "r1", "r2", "r3", "r4", "r5",
                            "r6", "r7", "r8", "r9", "r10")

# load the recommendations from the recommender output object into a data frame
for (i in 1:length(b_pred@items)){
  
  # get the recommended artists for the user
  tmp_recs <- as.vector(b_pred@items[[i]])
  
  # get the length of rec vector for the user
  num_trecs <- length(tmp_recs)

  # get list of unique user's artists from original data
  user_arts <- unique(subset(last_sm, userID == user_ids[i])$artistID)
  
  # eliminate artist that are already in user's playlist history
  new_recs <- tmp_recs[!(tmp_recs %in% user_arts) ]

  # get the length of new_rec vector
  num_newrecs <- length(new_recs)
  
  # if too few recommendations generated, sample 10 at random from the top815
  if(num_newrecs < 10) {
    new_recs <- sample(top_815$artistID[!(top_815$artistID %in% user_arts)], 10)
    
  }
  
  # if too few recs to implement strategy, just use the first 10
  if (num_newrecs < 13) {
    topten <- new_recs[1:10]
  } else {
    # randomly select 7 of the top 10 remaining recommendations
    t_seven <- sample(new_recs[1:10], 7)
    
    # then randomly select 3 of the remaining recommendations
    t_three <- sample(new_recs[11:length(new_recs)], 3)
    
    # merge the two lists of artist ID's
    topten <- c(t_seven, t_three)
  } # end if else
  
  # scramble the top 10 so that they are randomly ordered
  topten <- sample(topten, 10)
  
  # add recs to data frame
  user_tenrecs[i,2:11 ] <- topten
  
} # end for loop
```


_____

### Display a List of 10 Recommended Artists for a Given User

we can randomly select an User ID and display a list of recommended similar artists:

```{r}
# randomly select a user
user <- sample(user_ids, 1)

# fetch their recommendations
urecs <- sort(as.vector(subset(user_tenrecs, userID == user)[2:11]) )

# create list of artist names from artist ID's in list
rec_names <- subset(lfm_art, id %in% urecs)$name

kable(rec_names, col.names = "Artists You Might Enjoy")
```


_____

# Recommend Similar Artists via an Artist Similarity Matrix

Create a similarity matrix using cosine distance as metric of similarity

```{r}
# calculate artist similarity matrix
art_sim <- similarity(as(bin_agmat, "binaryRatingMatrix"), method = "cosine",
                     which = "users")

# convert to an R matrix
art_sim <- as(art_sim, "matrix")

# round to 3 digit precision
art_sim[][] <- round(art_sim[][],3)

# # name rows + cols according to artistID for easy retrieval
colnames(art_sim) <- ag_artistID
rownames(art_sim) <- ag_artistID

```


_____

### Generate Top N Similar Artist list for a Given ArtistID

we can randomly select an Artist ID and display a list of recommended similar artists:

```{r}
# set number of similar artists to recommend
n_recommended <- 5

# randomly select a user
artist <- sample(ag_artistID, 1)

# get name of artist from artist list
a_name <- lfm_art[lfm_art$id == artist,]$name

# fetch their recommendations: this returns a named vector sorted by similarity
# the names of the items are the artist IDs
arecs <- sort(art_sim[as.character(artist),], decreasing = TRUE)[1:n_recommended]

# extract the artist IDs and convert to numeric
arecs_IDs <- as.numeric(names(arecs))

# create list of artist names from artist ID's in list
arec_names <- lfm_art[lfm_art$id %in% arecs_IDs,]$name

# create a heading for the list of similar artists
table_head <- sprintf("Artists Similar to %s", a_name)

# display the list of similar artists
kable(arec_names, col.names = table_head)
```


_____

# Generate a Top N Artist List by Genre

Need to use the original artist-genre matrix:

```{r}
# this is only here for random number generation: delete in production mode
set.seed(42)

# set rownames = artistID's for easy retrieval - DON'T NEED THIS LINE OF CODE IN SHINY
rownames(ag_mat) <- ag_artistID

# extract the genre tagIDs from matrix and convert to numeric
tagIDs <- as.numeric(colnames(ag_mat))

# set number of artists to recommend
n_recommended <- 5

# randomly select a genre
tagID <- sample(tagIDs, 1)

# get name of genre from tagID list
g_name <- lfm_tags[lfm_tags$tagID == tagID,]$tagValue

# fetch the top N artists:
# the names of the items are the artist IDs
g_arecs <- sort(ag_mat[,as.character(tagID)], decreasing = TRUE)[1:n_recommended]

# extract the artist IDs and convert to numeric
g_arecs_IDs <- as.numeric(names(g_arecs))

# create list of artist names from artist ID's in list
g_arec_names <- lfm_art[lfm_art$id %in% g_arecs_IDs,]$name

# create a heading for the list of similar artists
table_head <- sprintf("Top Artists in %s genre:", g_name)

# display the list of similar artists
kable(g_arec_names, col.names = table_head)

```

_____

# Save R objects to files for use in Shiny App

Save user_tenrecs object to an RDA file for use in Shiny App:

```{r, eval = FALSE}
# save an R object to a file for future use
save(user_tenrecs, file = "c:/data/643/user_tenrecs.RData")

# delete the file from memory
rm(user_tenrecs)

# reload delete object into memory
load(file = "c:/data/643/user_tenrecs.RData")
```

Save artist-genre matrix to a file for use in Shiny App:

```{r, eval = FALSE}
# save an R object to a file for future use
save(ag_mat, file = "c:/data/643/ag_mat.RData")

# delete the file from memory
rm(ag_mat)

# reload delete object into memory
load(file = "c:/data/643/ag_mat.RData")
```


Save artist similarity matrix object to a file for use in Shiny App:

```{r, eval = FALSE}
# save an R object to a file for future use
save(art_sim, file = "c:/data/643/art_sim.RData")

# delete the file from memory
rm(art_sim)

# reload delete object into memory
load(file = "c:/data/643/art_sim.RData")
```


Save user-artist matrix (reduced per code herein to top 815 artists). We will need the user-artist matrix if we want to be able to allow the user to browse a list of artists they've previously listened to.

```{r, eval = FALSE}
# save an R object to a file for future use
save(last_sm, file = "c:/data/643/last_sm.RData")

# delete the file from memory
rm(last_sm)

# reload delete object into memory
load(file = "c:/data/643/last_sm.RData")
```